---
title: 'Exploring Data Frames & Data frame Manipulation with dplyr '
teaching: 10
exercises: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

# Load libraries ----------------------------------------------------------
# Package names
packages <- c("tidyverse", "here")

# Packages loading
invisible(lapply(packages, library, character.only = TRUE))

```

:::::::::::::::::::::::::::::::::::::: questions 

- What is a data frame?
- How can I read data in R?
- How can I get basic summary information about my data set?
- How can I select specific rows and/or columns from a data frame?
- How can I combine multiple commands into a single command?
- How can I create new columns or remove existing columns from a data frame?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

After completing this episode, participants should be able to…

- Describe what a data frame is.
- Load external data from a .csv file into a data frame.
- Summarize the contents of a data frame.
- Select certain columns in a data frame with the dplyr function select.
- Select certain rows in a data frame according to filtering conditions with the dplyr function filter.
- Link the output of one dplyr function to the input of another function with the ‘pipe’ operator |>.
- Add new columns to a data frame that are functions of existing columns with mutate.
- Use the split-apply-combine concept for data analysis.
- Use summarize, group_by, and count to split a data frame into groups of observations, apply a summary statistics for each group, and then combine the results.

::::::::::::::::::::::::::::::::::::::::::::::::


# Exploring Data frames

Now we turn to the bread-and-butter of working with `R`: working with tabular data. In `R` data are stored in a data structure called **data frames**.  

A data frame is a representation of data in the format of a **table** where the columns are **vectors** that all have the **same length**. 


Because columns are vectors, each column must contain a **single type of data** (e.g., characters, numeric, factors). 
For example, here is a figure depicting a data frame comprising a numeric, a character, and a logical vector.

![A data frame](fig/data-frame.svg)
<br><font size="3">*Source*: [Data Carpentry R for Social Scientists ](https://datacarpentry.org/r-socialsci/02-starting-with-data/index.html#what-are-data-frames-and-tibbles)</font>


## Reading data

`read.csv()` is a function used to read coma separated data files (`.csv` format)). There are other functions for files separated with other delimiters. 
We're gonna read in the `gapminder` data set with information about countries' size, GDP and average life expectancy in different years.

```{r reading-data}
gapminder <- read.csv(here("data", "gapminder_data.csv"))

```

## Exploring dataset
Let’s investigate the `gapminder` data frame a bit; the first thing we should always do is check out what the data looks like.

It is important to see if all the variables (columns) have the data type that we require. For instance, a column might have numbers stored as characters, which would not allow us to make calculations with those numbers.

```{r inspecting-data-str}
str(gapminder) 

```

We can see that the `gapminder` object is a data.frame with `r nrow(gapminder)` observations (rows) and `r ncol(gapminder)` variables (columns). 

In each line after a `$` sign, we see the name of each column, its type and first few values. 


### First look at the dataset
There are multiple ways to explore a data set. Here are just a few examples:


```{r examine-data}

head(gapminder) # shows first 6  rows of the data set

summary(gapminder) # basic statistical information about each column.
# Information format differes by data type.

nrow(gapminder) # returns number of rows in a dataset

ncol(gapminder) # returns number of columns in a dataset

```

### Dollar sign ($)

When you're analyzing a data set, you often need to access its specific columns.

One handy way to access a column is using it's name and a dollar sign `$`: 
```{r subset-dollar-sign}

# This notation means: From dataset gapminder, give me column country. You can
# see that the column accessed in this way is just a vector of characters.
country_vec <- gapminder$country

head(country_vec)
```

Now you can explore distinct values from a vector with the unique() function:
```{r subset-unique}
head(unique(country_vec), 10)

```
Note that the calling a column with a `$` sign will return a *vector* - it's not a data frame anymore.


# Data frame Manipulation with dplyr

## Select
Let's start manipulating the data. 

First, we will adapt our data set, by keeping only the columns we're interested in, using the `select()` function from the `dplyr` package:

```{r dplyr-select}
year_country_gdp <- select(gapminder, year, country, gdpPercap)

head(year_country_gdp)

```

## Pipe
Now, this is not the most common notation when working with `dplyr` package.
`R` offers an operator `|>` called a pipe, which allows you to build up complicated commands in a readable way.


::: callout

# The pipe

The `|>` operator, also called the "native pipe", was introduced in `R` version 4.1.0. 
Before that, the `%>%` operator from the `magrittr` package was widely used. 
The two pipes work in similar ways. 
The main difference is that you don't need to load any packages to have the native pipe available.

:::

The `select()` statement with pipe would look like that:

```{r dplyr-pipe}

year_country_gdp <- gapminder |>
  select(year, country, gdpPercap)

head(year_country_gdp)

```

First we define data set, then - with the use of pipe we pass it on to the `select()` function. 
This way we can chain multiple functions together, which we will be doing now. 

## Filter

We already know how to select only the needed columns. 
But now, we also want to filter the rows of our data set on certain conditions
with the `filter()` function. Instead of doing it in separate steps, we can do it all together. 

In the `gapminder` data set, we want to see the results from outside of Europe for the 21st century. 

```{r filter-data}
year_country_gdp_euro <- gapminder |>
  filter(continent != "Europe" & year >= 2000) |>
  select(year, country, gdpPercap)
# '&' operator (AND) - both conditions must be met

head(year_country_gdp_euro)
```

Let's now focus only on North American countries  
```{r filter-data-or}
year_gdp_namerica <- year_country_gdp_euro %>%
  filter(country == "Canada" |  country == "Mexico" | country == "United States") 

# '|' operator (OR) - at least one of the conditions must be met

head(year_gdp_namerica)
```

::: challenge

##  Challenge: filtered data frame

Write a single command (which can span multiple lines and includes pipes) 
that will produce a data frame that has the values for **life expectancy**, 
**country** and **year**, only for **EurAsia**. 

How many rows does your data frame have and why? 

::: solution

```{r ex5, class.source="bg-info"}
year_country_gdp_eurasia <- gapminder |>
  filter(continent == "Europe" | continent == "Asia") |>
  select(year, country, gdpPercap)
# '|' operator (OR) - one of the conditions must be met

nrow(year_country_gdp_eurasia)
``` 

:::

:::

## Group and summarize
So far, we have provided summary statistics on the whole dataset, selected columns, and filtered the observations. But often instead of doing that, we would like to know statistics about all of the continents, presented by group.

```{r dplyr-group}
gapminder |> # select the dataset
  group_by(continent) |> # group by continent
  summarize(avg_gdpPercap = mean(gdpPercap)) # create basic stats

```

::: challenge

## Challenge: longest and shortest life expectancy

Calculate the average life expectancy per country. Which country has the longest average life expectancy and which has the shortest average life expectancy?

<strong>Hint</strong> Use `max()`  and `min()` functions to find minimum and maximum.

::: solution

```{r ex6 , class.source="bg-info"}
gapminder |>
  group_by(country) |>
  summarize(avg_lifeExp = mean(lifeExp)) |>
  filter(avg_lifeExp == min(avg_lifeExp) |
    avg_lifeExp == max(avg_lifeExp))
```

:::

:::

### Multiple groups and summary variables
You can also group by multiple columns:

```{r dplyr-group-multi}

gapminder |>
  group_by(continent, year) |>
  summarize(avg_gdpPercap = mean(gdpPercap))

```

On top of this, you can also make multiple summaries of those groups:
```{r dplyr-summ}
gdp_pop_bycontinents_byyear <- gapminder |>
  group_by(continent, year) |>
  summarize(
    avg_gdpPercap = mean(gdpPercap),
    sd_gdpPercap = sd(gdpPercap),
    avg_pop = mean(pop),
    sd_pop = sd(pop),
    n_obs = n()
  )

```

## Frequencies

If you need only a number of observations per group, you can use the `count()` function
```{r dplyr-count}

gapminder |>
  count(continent)

```
 

## Mutate

Frequently you’ll want to create new columns based on the values in existing columns. For example, instead of only having the GDP per capita, we might want to create a new GDP variable and convert its units into Billions. For this, we’ll use `mutate()`.

```{r dplyr-mutate}
gapminder_gdp <- gapminder |>
  mutate(gdpBillion = gdpPercap * pop / 10^9)

head(gapminder_gdp)

```


::::::::::::::::::::::::::::::::::::: keypoints 

- We can use the `select()` and `filter()` functions to select certain columns in a data frame and to subset it based a specific conditions.
- With `mutate()`, we can create new columns in a data frame with values based on existing columns.
- By combining `group_by()` and `summarize()` in a pipe (`|>`) chain,  we can generate summary statistics for each group in a data frame.

::::::::::::::::::::::::::::::::::::::::::::::::

