---
title: 'Explore and plot by vector layer attributes'
teaching: 20
exercises: 10
---

```{r setup, include=FALSE}
library(tidyverse)  # tools for wrangling, reshaping and visualizing data
library(here)       # managing paths
library(sf)         # work with spatial vector data

boundary_Delft <- st_read("data/delft-boundary.shp")
lines_Delft <- st_read("data/delft-streets.shp")
point_Delft <- st_read("data/delft-leisure.shp")
```


:::::::::::::::::::::::::::::::::::::: questions 

- How can I examine the attributes of a vector layer?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

After completing this episode, participants should be able to…

- Query attributes of a vector object.

- Subset vector objects using specific attribute values.

- Plot a vector feature, colored by unique attribute values.

::::::::::::::::::::::::::::::::::::::::::::::::

# Query Vector Feature Metadata

Let's have a look at the content of the loaded data, starting with `lines_Delft`. In essence, an `"sf"` object is a data.frame with a "sticky" geometry column and some extra metadata, like the CRS, extent and geometry type we examined earlier.

```{r}
lines_Delft
```

This means that we can examine and manipulate them as data frames. For instance, we can look at the number of variables (columns in a data frame) with `ncol()`.

```{r}
ncol(lines_Delft)
```

In the case of `point_Delft` those columns are `"osm_id"`, `"highway"` and `"geometry"`. We can check the names of the columns with the base R function `names()`.

```{r}
names(lines_Delft)
```

::: callout

Note that in R the geometry is just another column and counts towards the number
returned by `ncol()`. This is different from GIS software with graphical user 
interfaces, where the geometry is displayed in a viewport not as a column in the 
attribute table.

:::

We can also preview the content of the object by looking at the first 6 rows with the `head()` function, which is in the case of an `sf` object the same as examining the object directly.

```{r}
head(lines_Delft)
```


## Explore values within one attribute

Using the `$` operator, we can examine the content of a single field of our lines feature. 

We can see the contents of the `highway` field of our lines feature:

```{r}
head(lines_Delft$highway, 10)
```

To see only unique values within the `highway` field, we can use the `unique()` function. This function extracts all possible values of a character variable. 

```{r}
unique(lines_Delft$highway)
```

:::::::::::::::::::::::: callout 

R is also able to handle categorical variables called factors. With factors, we can use the `levels()` function to show unique values. To examine unique values of the `highway` variable this way, we have to first transform it into a factor with the `factor()` function:

```{r}
levels(factor(lines_Delft$highway))
```

::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge 

## Challenge: Attributes for different spatial classes
<!-- 3 minutes -->

Explore the attributes associated with the `point_Delft` spatial object.

1. How many attributes does it have?
2. What types of leisure points do the points represent? Give three examples.
3. Which of the following is NOT an attribute of the `point_Delft` data object?

  A) location B) leisure C) osm_id

:::::::::::::::::::::::: solution 

1. To find the number of attributes, we use the `ncol()` function:

```{r}
ncol(point_Delft)
```

2. The types of leisure point are in the column named `leisure`.

Using the `head()` function which displays 6 rows by default, we only see two values and `NA`s. 

```{r}
head(point_Delft)
```

We can increase the number of rows with the n argument (e.g., `head(n = 10)` to show 10 rows) until we see at least three distinct values in the leisure column. Note that printing an `sf` object will also display the first 10 rows.

```{r}
head(point_Delft, 10)  # you might be lucky to see three distinct values
# point_Delft
```

However, this is not a good approach as the first rows might still have many `NA`s and three distinct values might still not be present in the first `n` rows of the data frame. To remove `NA`s, we can use the function `na.omit()` on the leisure column to remove `NA`s completely. Note that we use the `$` operator to examine the content of a single variable.

```{r}
head(na.omit(point_Delft$leisure))  # this is better
```

To show only unique values, we can use the `unique()` function to only see the first occurrence of each distinct value (note that `NA` will still be included once).

```{r}
head(unique(point_Delft$leisure), n = 3)   # this is even better
```

3. To see a list of all attribute names, we can use the `names()` function.

```{r}
names(point_Delft)
```

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::


## Subset features

We can use the `filter()` function to select a subset of features from a spatial object, just like with data frames. Let's select only cycleways from our street data. 

```{r}
cycleway_Delft <- lines_Delft %>%  
  filter(highway == "cycleway")
```

Our subsetting operation reduces the number of features from `r nrow(lines_Delft)` to `r nrow(cycleway_Delft)`.

```{r}
nrow(lines_Delft)
nrow(cycleway_Delft)
```

This can be useful, for instance, to calculate the total length of cycleways.

```{r}
cycleway_Delft <- cycleway_Delft %>% 
  mutate(length = st_length(.))

cycleway_Delft %>%
  summarise(total_length = sum(length))
```

Now we can plot only the cycleways.

```{r fig.cap="Map of cycleways in Delft."}
ggplot(data = cycleway_Delft) +
  geom_sf() +
  labs(title = "Slow mobility network in Delft", subtitle = "Cycleways") +
  coord_sf(datum = st_crs(28992))
```

::: challenge
<!-- 7 minutes -->

1. Create a new object that only contains the motorways in Delft. 
2. How many features does the new object have?
3. What is the total length of motorways?
4. Plot the motorways.
5. Extra: follow the same steps with pedestrian streets. 

::: solution

1. To create the new object, we first need to see which value of the `highway` column holds motorways. There is a value called `motorway`.

```{r}
unique(lines_Delft$highway)
```
We extract only the features with the value `motorway`.

```{r}
motorway_Delft <- lines_Delft %>% 
  filter(highway == "motorway")

motorway_Delft
```

2. There are `r nrow(motorway_Delft)` features with the value `motorway`.

```{r}
motorway_Delft_length <- motorway_Delft %>% 
  mutate(length = st_length(.)) %>% 
  select(everything(), geometry) %>%
  summarise(total_length = sum(length))
```

3. The total length of motorways is `r motorway_Delft_length$total_length`.

```{r}
nrow(motorway_Delft)
```

4. Plot the motorways.

```{r}
ggplot(data = motorway_Delft) + 
  geom_sf(linewidth = 1.5) +
  labs(title = "Fast mobility network", subtitle = "Motorways") + 
  coord_sf(datum = st_crs(28992))
```

5. Follow the same steps with pedestrian streets.

```{r}
pedestrian_Delft <- lines_Delft %>% 
  filter(highway == "pedestrian")

pedestrian_Delft %>% 
  mutate(length = st_length(.)) %>% 
  select(everything(), geometry) %>%
  summarise(total_length = sum(length))

nrow(pedestrian_Delft)
```

```{r}
ggplot() +
  geom_sf(data = pedestrian_Delft) +
  labs(title = "Slow mobility network", subtitle = "Pedestrian") + 
  coord_sf(datum = st_crs(28992))
```

:::

:::

## Customize plots

Let's say that we want to color different road types with different colors and that we want to determine those colors.

```{r}
unique(lines_Delft$highway)
```

If we look at all the unique values of the highway field of our street network we see more than 20 values. Let's focus on a subset of four values to illustrate the use of distinct colors. We use a piped expression in which we only filter the rows of our data frame that have one of the four given values `"motorway"`, `"primary"`, `"secondary"`, and `"cycleway"`. We also make sure that the highway column is a factor column.

```{r}
road_types <- c("motorway", "primary", "secondary", "cycleway")

lines_Delft_selection <- lines_Delft %>% 
  filter(highway %in% road_types) %>% 
  mutate(highway = factor(highway, levels = road_types))
```

Next we define the four colors we want to use, one for each type of road in our vector object. Note that in R you can use named colors like `"blue"`, `"green"`, `"navy"`, and `"purple"`. A full list of named colors can be listed with the `colors()` function.

```{r}
road_colors <- c("blue", "green", "navy", "purple")
```

We can use the defined color palette in ggplot.

```{r}
ggplot(data = lines_Delft_selection) +
  geom_sf(aes(color = highway)) + 
  scale_color_manual(values = road_colors) +
  labs(color = 'Road Type',
       title = "Road network of Delft", 
       subtitle = "Roads & Cycleways") + 
  coord_sf(datum = st_crs(28992))
```

## Adjust line width

Earlier we adjusted the line width universally. We can also adjust line widths for every factor level. Note that in this case the `size` argument, like the `color` argument, are within the `aes()` mapping function. This means that the values of that visual property will be mapped from a variable of the object that is being plotted.

```{r}
line_widths <- c(1, 0.75, 0.5, 0.25)
```

```{r}
ggplot(data = lines_Delft_selection) +
  geom_sf(aes(color = highway, linewidth = highway)) +
  scale_color_manual(values = road_colors) +
  labs(color = 'Road Type',
       linewidth = 'Road Type',
       title = "Mobility network of Delft",
       subtitle = "Roads & Cycleways") +
  scale_linewidth_manual(values = line_widths) +
  coord_sf(datum = st_crs(28992))
```

::: challenge

# Plot line width by attribute

In the example above, we set the line widths to be 1, 0.75, 0.5, and 0.25. In our case line thicknesses are consistent with the hierarchy of the selected road types, but in some cases we might want to show a different hierarchy.

Let’s create another plot where we show the different line types with the following thicknesses:

- motorways linewidth = 0.25
- primary linewidth = 0.75
- secondary linewidth =  0.5
- cycleway linewidth = 1

::: solution

```{r}
levels(factor(lines_Delft_selection$highway))
```

```{r}
line_width <- c(0.25, 0.75, 0.5, 1)
```

```{r}
ggplot(data = lines_Delft_selection) +
  geom_sf(aes(linewidth = highway)) +
  scale_linewidth_manual(values = line_width) +
  labs(title = "Mobility network of Delft",
       subtitle = "Roads & Cycleways - Line width varies") + 
  coord_sf(datum = st_crs(28992))
```

:::

:::


## Add plot legend

Let’s add a legend to our plot. We will use the `road_colors` object that we created above to color the legend. We can customize the appearance of our legend by manually setting different parameters.

```{r fig.cap="Mobility network in Delft using thicker lines than the previous example."}
ggplot(data = lines_Delft_selection) + 
  geom_sf(aes(color = highway), linewidth = 1.5) +
  scale_color_manual(values = road_colors) +
  labs(color = 'Road Type') + 
  labs(title = "Mobility network of Delft", 
       subtitle = "Roads & Cycleways - Default Legend") + 
  coord_sf(datum = st_crs(28992))
```

```{r fig.cap="Map of the mobility network in Delft with large-font and border around the legend."}
ggplot(data = lines_Delft_selection) + 
  geom_sf(aes(color = highway), linewidth = 1.5) +
  scale_color_manual(values = road_colors) + 
  labs(color = 'Road Type') +
  theme(legend.text = element_text(size = 20), 
        legend.box.background = element_rect(size = 1)) + 
  labs(title = "Mobility network of Delft", 
       subtitle = "Roads & Cycleways - Modified Legend") +
  coord_sf(datum = st_crs(28992))
```

```{r fig.cap="Map of the mobility network in Delft using a different color palette."}
new_colors <- c("springgreen", "blue", "magenta", "orange")

ggplot(data = lines_Delft_selection) + 
  geom_sf(aes(color = highway), linewidth = 1.5) + 
  scale_color_manual(values = new_colors) +
  labs(color = 'Road Type') +
  theme(legend.text = element_text(size = 20), 
        legend.box.background = element_rect(size = 1)) + 
  labs(title = "Mobility network of Delft", 
       subtitle = "Roads & Cycleways - Modified Legend") +
  coord_sf(datum = st_crs(28992))
```

::: challenge

# Plot lines by attributes
<!-- 5 minutes -->

Create a plot that emphasizes only roads where bicycles are allowed. To emphasize this, make the lines where bicycles are not allowed THINNER than the roads where bicycles are allowed. Be sure to add a title and legend to your map. You might consider a color palette that has all bike-friendly roads displayed in a bright color. All other lines can be black.

::: solution

```{r}
class(lines_Delft_selection$highway)
```

```{r}
levels(factor(lines_Delft$highway))
```

```{r}
# First, create a data frame with only those roads where bicycles are allowed
lines_Delft_bicycle <- lines_Delft %>% 
  filter(highway == "cycleway")

# Next, visualise using ggplot
ggplot(data = lines_Delft) +
  geom_sf() +
  geom_sf(data = lines_Delft_bicycle, aes(color = highway), linewidth = 1) +
  scale_color_manual(values = "magenta") +
  labs(title = "Mobility network in Delft", subtitle = "Roads dedicated to Bikes") +
  coord_sf(datum = st_crs(28992))
```

:::

:::

::: challenge

# Plot polygon by attribute
<!-- 5 minutes -->

Create a map of the municipal boundaries in the Netherlands using the data located in your data folder: `nl-gemeenten.shp`. Apply a line color to each state using its region value. Add a legend.

::: solution

```{r}
municipal_boundaries_NL <- st_read(here("episodes", "data", "nl-gemeenten.shp"))
```

```{r}
str(municipal_boundaries_NL)
levels(factor(municipal_boundaries_NL$ligtInPr_1))
```

```{r}
ggplot(data = municipal_boundaries_NL) +
  geom_sf(aes(color = ligtInPr_1), linewidth = 1) +
  labs(title = "Contiguous NL Municipal Boundaries") + 
  coord_sf(datum = st_crs(28992))
```

:::

:::


::::::::::::::::::::::::::::::::::::: keypoints 

- Spatial objects in `sf` are similar to standard data frames and can be manipulated using the same functions.

- Almost any feature of a plot can be customized using the various functions and options in the `ggplot2` package.

::::::::::::::::::::::::::::::::::::::::::::::::

