---
title: 'Plot multiple shapefiles'
teaching: 40
exercises: 20
---

```{r setup, include=FALSE}
library(tidyverse)  # tools for wrangling, reshaping and visualizing data
library(here)       # managing paths
library(sf)         # work with spatial vector data

boundary_Delft <- st_read("data/delft-boundary.shp")
lines_Delft <- st_read("data/delft-streets.shp")
point_Delft <- st_read("data/delft-leisure.shp")

road_types <- c("motorway", "primary", "secondary", "cycleway")
road_colors <- c("blue", "green", "navy", "purple")

lines_Delft_selection <- lines_Delft %>% 
  filter(highway %in% road_types) %>% 
  mutate(highway = factor(highway, levels = road_types))
```

:::::::::::::::::::::::::::::::::::::: questions 

- How can I create map compositions with custom legends using ggplot?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

After completing this episode, participants should be able to…

- Plot multiple vector layers in the same plot.
- Apply custom symbols to spatial objects in a plot.

::::::::::::::::::::::::::::::::::::::::::::::::

This episode builds upon the [previous episode](../episodes/10-explore-and-plot-by-vector-layer-attributes.Rmd) to work with vector layers in R and explore how to plot multiple vector layers.



## Load the data

To work with vector data in R, we can use the `sf` library. The `terra` package also allows us to explore metadata using similar commands for both raster and vector files. Make sure that you have these packages loaded.

We will continue to work with the three ESRI shapefiles that we loaded in the [Open and Plot Vector Layers](../episodes/09-open-and-plot-vector-layers.Rmd) episode.



## Plotting Multiple Vector Layers

So far we learned how to plot information from a single shapefile and do some plot customization. What if we want to create a more complex plot with many shapefiles and unique symbols that need to be represented clearly in a legend?

We will create a plot that combines our leisure locations (`point_Delft`), municipal boundary (`boundary_Delft`) and streets (`lines_Delft`) spatial objects. We will need to build a custom legend as well.

To begin, we will create a plot with the site boundary as the first layer. Then layer the leisure locations and street data on top using `+`.

```{r}
ggplot() + 
  geom_sf(data = boundary_Delft, fill = "lightgrey", color = "lightgrey") +
  geom_sf(data = lines_Delft_selection, aes(color = highway), size = 1) +
  geom_sf(data = point_Delft) +
  labs(title = "Mobility network of Delft") + 
  coord_sf(datum = st_crs(28992))
```

Next, let’s build a custom legend using the symbology (the colors and symbols) that we used to create the plot above. For example, it might be good if the lines were symbolized as lines. In the previous episode, you may have noticed that the default legend behavior for `geom_sf` is to draw a ‘patch’ for each legend entry. If you want the legend to draw lines or points, you need to add an instruction to the `geom_sf` call - in this case, `show.legend = 'line'`.

```{r}
leisure_colors <- rainbow(15)
point_Delft$leisure <- factor(point_Delft$leisure)
ggplot() + 
  geom_sf(data = boundary_Delft, fill = "lightgrey", color = "lightgrey") +
  geom_sf(data = lines_Delft_selection, aes(color = highway), size = 1) + 
  geom_sf(data = point_Delft, aes(fill = leisure), shape = 21) +
  scale_color_manual(values = road_colors, name = "Road Type") + 
  scale_fill_manual(values = leisure_colors, name = "Lesiure Location") + 
  labs(title = "Mobility network and leisure in Delft") + 
  coord_sf(datum = st_crs(28992))
```

```{r}
ggplot() +
  geom_sf(data = boundary_Delft, fill = "lightgrey", color = "lightgrey") +
  geom_sf(data = lines_Delft_selection, aes(color = highway), size = 1) +
  geom_sf(data = point_Delft, aes(fill = leisure), shape = 22) +
  scale_color_manual(values = road_colors, name = "Line Type") +
  scale_fill_manual(values = leisure_colors, name = "Leisure Location") +
  labs(title = "Mobility network and leisure in Delft") + 
  coord_sf(datum = st_crs(28992))
```

We notice that there are quite some playgrounds in the residential parts of Delft, whereas on campus there is a concentration of picnic tables. So that is what our next challenge is about.

### Challenge 7 (5 minutes)

Create a map of leisure locations only including `playground` and `picnic_table`, with each point colored by the leisure type. Overlay this layer on top of the `lines_Delft` layer (the streets). Create a custom legend that applies line symbols to lines and point symbols to the points.

Modify the plot above. Tell R to plot each point, using a different symbol of shape value.

```{r}
leisure_locations_selection <- st_read("data/delft-leisure.shp") %>% 
  filter(leisure %in% c("playground", "picnic_table"))
```

```{r}
levels(factor(leisure_locations_selection$leisure))
```

```{r}
blue_orange <- c("cornflowerblue", "darkorange")
```

```{r}
ggplot() + 
  geom_sf(data = lines_Delft_selection, aes(color = highway)) + 
  geom_sf(data = leisure_locations_selection, aes(fill = leisure), 
          shape = 21, show.legend = 'point') + 
  scale_color_manual(name = "Line Type", values = road_colors,
     guide = guide_legend(override.aes = list(linetype = "solid", shape = NA))) + 
  scale_fill_manual(name = "Soil Type", values = blue_orange,
     guide = guide_legend(override.aes = list(linetype = "blank", shape = 21, colour = NA))) + 
  labs(title = "Traffic and leisure") + 
  coord_sf(datum = st_crs(28992))
```

```{r}
ggplot() + 
  geom_sf(data = lines_Delft_selection, aes(color = highway), size = 1) + 
  geom_sf(data = leisure_locations_selection, aes(fill = leisure, shape = leisure), size = 2) + 
  scale_shape_manual(name = "Leisure Type", values = c(21, 22)) +
  scale_color_manual(name = "Line Type", values = road_colors) + 
  scale_fill_manual(name = "Leisure Type", values = rainbow(15),
     guide = guide_legend(override.aes = list(linetype = "blank", shape = c(21, 22),
     color = "black"))) + 
  labs(title = "Road network and leisure") + 
  coord_sf(datum = st_crs(28992))
```

<!-- End of challenge -->






::::::::::::::::::::::::::::::::::::: keypoints 

- Use the `+` operator to add multiple layers to a ggplot.
- A plot can be a combination of multiple vector layers.
- Use the `show.legend` argument to set legend symbol types.
- Use the `scale_fill_manual()` function to set legend colors.

::::::::::::::::::::::::::::::::::::::::::::::::

