---
title: 'Handling Spatial Projections & CRS'
teaching: 40
exercises: 20
---

```{r setup, include=FALSE}
library(tidyverse)  # tools for wrangling, reshaping and visualizing data
library(here)       # managing paths
library(sf)         # work with spatial vector data
library(terra)      # work with spatial vector and raster metadata

leisure_locations_selection <- st_read("data/delft-leisure.shp") %>% 
  filter(leisure %in% c("playground", "picnic_table"))
```

:::::::::::::::::::::::::::::::::::::: questions 

- What do I do when vector data don’t line up?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives


After completing this episode, participants should be able to…

- Plot vector objects with different CRSs in the same plot.

::::::::::::::::::::::::::::::::::::::::::::::::

## Working with spatial data from different sources

```{r}
municipal_boundary_NL <- st_read("data/nl-gemeenten.shp")
```

```{r}
ggplot() +
  geom_sf(data = municipal_boundary_NL) +
  labs(title = "Map of Contiguous NL Municipal Boundaries") +
  coord_sf(datum = st_crs(28992))
```

We can add a country boundary layer to make it look nicer. If we specify a thicker line width using size = 2 for the country boundary layer, it will make our map pop!

```{r}
country_boundary_NL <- st_read("data/nl-boundary.shp")
```

```{r}
ggplot() +
  geom_sf(data = country_boundary_NL, color = "gray18", linewidth = 2) +
  geom_sf(data = municipal_boundary_NL, color = "gray40") +
  labs(title = "Map of Contiguous NL Municipal Boundaries") +
  coord_sf(datum = st_crs(28992))
```

```{r}
# st_crs(point_Delft)
```

```{r}
st_crs(municipal_boundary_NL)$epsg
```

```{r}
st_crs(country_boundary_NL)$epsg
```

```{r}
boundary_Delft <- st_read("data/delft-boundary.shp")
st_crs(boundary_Delft)$epsg
boundary_Delft <- st_transform(boundary_Delft, 28992)
```

```{r}
ggplot() +
  geom_sf(data = country_boundary_NL, linewidth = 2, color = "gray18") +
  geom_sf(data = municipal_boundary_NL, color = "gray40") +
  geom_sf(data = boundary_Delft, color = "purple", fill = "purple") +
  labs(title = "Map of Contiguous NL Municipal Boundaries") +
  coord_sf(datum = st_crs(28992))
```

::: challenge

# Plot multiple layers of spatial data
<!-- 5 minutes -->

Create a map of South Holland as follows:

1. Import and plot `nl-gemeenten.shp`. Adjust line width as necessary.
2. Layer the boundary of Delft onto the plot.
3. Add a title.
4. Add a legend that shows both the province boundaries (as a line) and the boundary of Delft (as a filled polygon).

::: solution

```{r}
boundary_ZH <- municipal_boundary_NL %>% 
  filter(ligtInPr_1 == "Zuid-Holland")
```

```{r}
ggplot() +
    geom_sf(data = boundary_ZH, aes(color ="color"), show.legend = "line") +
    scale_color_manual(name = "", labels = "Municipal Boundaries in South Holland", values = c("color" = "gray18")) +
    geom_sf(data = boundary_Delft, aes(shape = "shape"), color = "purple", fill = "purple") +
    scale_shape_manual(name = "", labels = "Municipality of Delft", values = c("shape" = 19)) +
    labs(title = "Delft location") +
    theme(legend.background = element_rect(color = NA)) +
    coord_sf(datum = st_crs(28992))
```

:::

:::



## Export a shapefile

To save a file, use the `st_write()` function from the `sf` package. Although `sf` guesses the driver needed for a specified output file name from its extension, this can be made explicitly via the `driver` argument. In our case `driver = "ESRI Shapefile"` ensures that the the output is correctly saved as a `.shp` file.

```{r, eval=FALSE}
st_write(leisure_locations_selection,
         "data/leisure_locations_selection.shp"), driver = "ESRI Shapefile")
```



::::::::::::::::::::::::::::::::::::: keypoints 

- `ggplot2` automatically converts all objects in a plot to the same CRS.
- Still be aware of the CRS and extent for each object.
- You can export an `sf` object to a shapefile with `st_write()`.

::::::::::::::::::::::::::::::::::::::::::::::::

