---
title: 'Handling Spatial Projections & CRS'
teaching: 25
exercises: 5
---

```{r setup, include=FALSE}
library(tidyverse)  # tools for wrangling, reshaping and visualizing data
library(here)       # managing paths
library(sf)         # work with spatial vector data

leisure_locations_selection <- st_read("data/delft-leisure.shp") %>%
  filter(leisure %in% c("playground", "picnic_table"))
```

:::::::::::::::::::::::::::::::::::::: questions 

- What do I do when vector data do not line up?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives


After completing this episode, participants should be able to…

- Plot vector objects with different CRSs in the same plot.

::::::::::::::::::::::::::::::::::::::::::::::::

## Working with spatial data from different sources

In this episode, we will work with a different dataset containing Dutch municipal 
boundaries. We start by reading the data and plotting it.

```{r minicipal-boundary}
municipal_boundary_NL <- st_read("data/nl-gemeenten.shp")
```

```{r plot-minicipal-boundary}
ggplot() +
  geom_sf(data = municipal_boundary_NL) +
  labs(title = "Map of Contiguous NL Municipal Boundaries") +
  coord_sf(datum = st_crs(28992))
```

We can add a country boundary layer to make it look nicer. If we specify a thicker line width using size = 2 for the country boundary layer, it will make our map pop! We read the country boundary from a different file.

```{r country-boundary}
country_boundary_NL <- st_read("data/nl-boundary.shp")
```

```{r plot-boundaries}
ggplot() +
  geom_sf(data = country_boundary_NL, 
          color = "gray18", 
          linewidth = 2) +
  geom_sf(data = municipal_boundary_NL, 
          color = "gray40") +
  labs(title = "Map of Contiguous NL Municipal Boundaries") +
  coord_sf(datum = st_crs(28992))
```

We confirm that the CRS of both boundaries is 28992.

```{r minicipal-boundary-crs}
st_crs(municipal_boundary_NL)$epsg
```

```{r country-boundary-crs}
st_crs(country_boundary_NL)$epsg
```

We read the municipal boundary of Delft and make sure that it is in the same CRS
as the country-level municipal boundaries and country boundary layers.

```{r municipal-boundary-crs-2}
boundary_Delft <- st_read("data/delft-boundary.shp")
st_crs(boundary_Delft)$epsg
boundary_Delft <- st_transform(boundary_Delft, 28992)
```

```{r plot-boundaries2}
ggplot() +
  geom_sf(data = country_boundary_NL, 
          linewidth = 2, 
          color = "gray18") +
  geom_sf(data = municipal_boundary_NL, 
          color = "gray40") +
  geom_sf(data = boundary_Delft, 
          color = "purple", 
          fill = "purple") +
  labs(title = "Map of Contiguous NL Municipal Boundaries") +
  coord_sf(datum = st_crs(28992))
```

::: challenge

# Challenge: Plot multiple layers of spatial data
<!-- 5 minutes -->

Create a map of South Holland as follows:

1. Import `nl-gemeenten.shp` and filter only the municipalities in South Holland.
2. Plot it and adjust line width as necessary.
3. Layer the boundary of Delft onto the plot.
4. Add a title.
5. Add a legend that shows both the municipal boundaries (as a line) and the boundary of Delft (as a filled polygon).

::: solution

```{r boundary-zh}
boundary_ZH <- municipal_boundary_NL %>% 
  filter(ligtInPr_1 == "Zuid-Holland")
```

```{r plot-boundary-zh}
ggplot() +
    geom_sf(data = boundary_ZH, 
            aes(color ="color"), 
            show.legend = "line") +
    scale_color_manual(name = "", 
                       labels = "Municipal Boundaries in South Holland", 
                       values = c("color" = "gray18")) +
    geom_sf(data = boundary_Delft, 
            aes(shape = "shape"), 
            color = "purple", 
            fill = "purple") +
    scale_shape_manual(name = "", 
                       labels = "Municipality of Delft", 
                       values = c("shape" = 19)
                       ) +
    labs(title = "Delft location") +
    theme(legend.background = element_rect(color = NA)) +
    coord_sf(datum = st_crs(28992))
```

:::

:::

::: callout

# Projecting layers

Note that `ggplot2` may reproject the layers on the fly for visualisation purposes, but for geoprocessing purposes, you still need to reproject the layers explicitly with `st_transform()`. This will become clear in a [later episode](../episodes/19-basic-gis-with-r-sf.Rmd) when we perform GIS operations.

:::

## Export a shapefile

To save a file, use the `st_write()` function from the `sf` package. Although `sf` guesses the driver needed for a specified output file name from its extension, this can be made explicitly via the `driver` argument. In our case `driver = "ESRI Shapefile"` ensures that the output is correctly saved as a `.shp` file.

```{r write-leisure-locations, eval=FALSE}
st_write(leisure_locations_selection,
  "data/leisure_locations_selection.shp",
  driver = "ESRI Shapefile"
)
```

::::::::::::::::::::::::::::::::::::: keypoints 

- `ggplot2` automatically converts all objects in a plot to the same CRS.
- For geoprocessing purposes, you still need to reproject the layers you use to the same CRS.
- You can export an `sf` object to a shapefile with `st_write()`.

::::::::::::::::::::::::::::::::::::::::::::::::

